---
title: "Social Distancing in Aberdeen"

output:
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
---

```{r setup, include=FALSE}
library(here)
library(tidyverse)
library(janitor)
library(sf)
library(flexdashboard)
library(mapview)
library(gt)
library(kableExtra)
```


```{r}
#load data and shapefiles 

simd_zones <- read_csv(here("data", "SIMD+2020v2+-+datazone+lookup.csv"))

simd_indicators <- read_csv(here("data", "SIMD2020v2_indicators.csv"))

datazone_sf <- st_read(here("data", "sc_dz_11.shp"), quiet = T)

occupations <- read_csv(here("data", "SNS Data Zone 2011 blk", "QS605SC.csv")) %>%
  clean_names()

#add columns with all occupations as percentages
occupations <-
occupations %>%
  mutate(
    across(
      a_agriculture_forestry_and_fishing:u_activities_of_extraterritorial_organisations_and_bodies,
      ~ .x/all_people_aged_16_to_74_in_employment *100, 
      .names = "perc_{col}"))

```


```{r}
#join SIMD files
#remove "%" and "*" and make numeric

simd <- full_join(simd_indicators, simd_zones, by = c("Data_Zone" = "DZ"))

simd <- full_join(simd, occupations, by = c("Data_Zone" = "x1"))

simd <- 
simd %>%
  mutate(across(everything(), ~ str_remove_all(., "%")),
         across(everything(), ~ str_remove_all(., "\\*")),
         across(Total_population:nocentralheat_rate, ~ as.numeric(.)),
         across(SIMD2020v2_Rank:Working_Age_Population, ~ as.numeric(.)),
         across(all_people_aged_16_to_74_in_employment:perc_u_activities_of_extraterritorial_organisations_and_bodies, ~ as.numeric(.)))

datazone_sf <-
datazone_sf %>% mutate(pop_per_km2 = TotPop2011/StdAreaKm2, 0)

#adds population density back in simd file from the shapefile
simd <-
datazone_sf %>%
  st_drop_geometry() %>%
  select(DataZone, pop_per_km2) %>%
  left_join(simd, ., by = c("Data_Zone" = "DataZone"))

# create index
#define essential jobs more narrowly????

grampian_index <-
simd %>%
  filter(HBname == "Grampian") %>%
  mutate(
    perc_essential = perc_c_manufacturing + perc_f_construction + perc_g_wholesale_and_retail_trade_repair_of_motor_vehicles_and_motorcycles + perc_i_accommodation_and_food_service_activities + perc_p_education + perc_q_human_health_and_social_work_activities,
         perc_working_age = Working_age_population/Total_population * 100) %>%
  select(Data_Zone, DZname, Council_area, Total_population, perc_working_age, SIMD2020v2_Rank, SIMD2020v2_Decile, overcrowded_rate, pop_per_km2, perc_essential)

grampian_index <-
grampian_index %>%
  mutate(scale_overcrowded = overcrowded_rate/max(overcrowded_rate) *10,
         scale_density = pop_per_km2/max(pop_per_km2) *10,
         scale_essential = perc_essential/max(perc_essential) *10,
         sum_scaled_score = scale_overcrowded + scale_density + scale_essential)

grampian_index <-
grampian_index %>%
  mutate(percentile_overcrowded = ntile(overcrowded_rate, 100),
         percentile_essential = ntile(perc_essential, 100),
         percentile_density = ntile(pop_per_km2, 100),
         sum_percentile_score = percentile_overcrowded + percentile_essential + percentile_density,
         sum_score_decile = ntile(sum_percentile_score, 10))

```


```{r}
#select Aberdeen City areas' data 
#join to shapefile

aberdeen <-
  grampian_index %>%
  filter(Council_area == "Aberdeen City")

aberdeen_data_zones <- pull(aberdeen, Data_Zone)

aberdeen_sf <- filter(datazone_sf, DataZone %in% aberdeen_data_zones)

aberdeen_sf <-
  merge(aberdeen_sf, aberdeen, by.x = "DataZone", by.y = "Data_Zone")

```



Social Distancing Score
=====================================

Row
-------------------------------------

###

```{r}
aberdeen_sf %>%
  mapview(
    map.types = "OpenStreetMap",
    zcol = "sum_score_decile", 
    label = aberdeen_sf$Name,
    layer.name = "",
    alpha.regions = 0.8
    )
```


###

```{r}
#areas with highest scores

aberdeen %>%
  select(Data_Zone, DZname, Total_population, sum_score_decile, sum_percentile_score) %>%
  filter(sum_score_decile == 10) %>%
  arrange(desc(sum_percentile_score)) %>%
  gt() %>%
  tab_header(
    title = "Areas with least ability to social distance", 
    subtitle = "Data zones scoring in the top 10% for difficulty social distancing") %>%
  cols_label(
    Data_Zone = "Data Zone",
    DZname = "Name", 
    Total_population = "Population", 
    sum_score_decile = "Distancing score",
    sum_percentile_score = "Total points")  %>%
  tab_options(container.height = 400)
```

Row
-------------------------------------

### Methods 

The Social Distancing Score ranks the regions in Aberdeen where people will find it most difficult to socially distance. 

The Social Distancing Score uses three measures: percent of working people employed in essential occupations, percent of people living in overcrowded homes, and high local population density. 

Aberdeen is divided into 283 areas (called data zones) of roughly equal population. Each area in Aberdeen has been given a score of 1 to 10 based on the populations' essential occupations, overcrowding, and population density. 

Essential occupations were defined as: manufacturing, construction, wholesale and retail trade, accomodation and food services, education, health and social care. 

For each measure, data zones were scored using their percentile rank in Aberdeen. The total score was then the decile of the summed percentiles. All scores are given in the Scores tab here. All raw data and code for calculating measures are given in the Sources tab here. 


Essential Jobs
=====================================

Row
-------------------------------------

###

```{r}
aberdeen_sf %>%
  mapview(
    map.types = "OpenStreetMap",
    zcol = "perc_essential", 
    label = aberdeen_sf$Name,
    layer.name = "",
    alpha.regions = 0.8
    )
```


###

```{r}
#areas with most people in essential occupations

aberdeen %>%
  select(Data_Zone, DZname, Total_population, perc_essential, sum_score_decile) %>%
  filter(perc_essential >= 65) %>%
  arrange(desc(perc_essential)) %>%
  gt() %>%
  fmt_number(columns = vars(perc_essential), decimals = 0) %>%
  tab_header(
    title = "Areas with most people in essential occupations", 
    subtitle = "Data zones with at least 65% of population in essential occupations") %>%
  cols_label(
    Data_Zone = "Data Zone",
    DZname = "Name", 
    Total_population = "Population", 
    sum_score_decile = "Distancing score",
    perc_essential = "Essential workers (%)")  %>%
  tab_options(container.height = 400)
```


Densest Areas
=====================================

Row
-------------------------------------

###

```{r}
aberdeen_sf %>%
  mapview(
    map.types = "OpenStreetMap",
    zcol = "pop_per_km2.x", 
    label = aberdeen_sf$Name,
    layer.name = "",
    alpha.regions = 0.8
    )
```


###

```{r}
#areas with most people per square kilometer

aberdeen %>%
  select(Data_Zone, DZname, Total_population, pop_per_km2, sum_score_decile) %>%
  filter(pop_per_km2 >= 10000) %>%
  arrange(desc(pop_per_km2)) %>%
  gt() %>%
  fmt_number(columns = vars(pop_per_km2), decimals = 0) %>%
  tab_header(
    title = "Areas with most people per km2", 
    subtitle = "Data zones with at least 10,000/km2") %>%
  cols_label(
    Data_Zone = "Data Zone",
    DZname = "Name", 
    Total_population = "Population", 
    sum_score_decile = "Distancing score",
    pop_per_km2 = "people/km2")  %>%
  tab_options(container.height = 400)
```


Overcrowded homes
=====================================

Row
-------------------------------------

###

```{r}
aberdeen_sf %>%
  mapview(
    map.types = "OpenStreetMap",
    zcol = "overcrowded_rate", 
    label = aberdeen_sf$Name,
    layer.name = "",
    alpha.regions = 0.8
    )
```


###

```{r}
#areas with most people living in overcrowded homes

aberdeen %>%
  select(Data_Zone, DZname, Total_population, overcrowded_rate, sum_score_decile) %>%
  filter(overcrowded_rate >= 25) %>%
  arrange(desc(overcrowded_rate)) %>%
  gt() %>%
  fmt_number(columns = vars(overcrowded_rate), decimals = 0) %>%
  tab_header(
    title = "Areas with most people living in overcrowded homes", 
    subtitle = "Data zones with at least 25% people in overcrowded homes") %>%
  cols_label(
    Data_Zone = "Data Zone",
    DZname = "Name", 
    Total_population = "Population", 
    sum_score_decile = "Distancing score",
    overcrowded_rate = "overcrowded (%)")  %>%
  tab_options(container.height = 400)
```

Scores
=====================================

Row
-------------------------------------

###


```{r}
#scores table

aberdeen %>%
  select(Data_Zone, DZname, Total_population, overcrowded_rate, percentile_overcrowded, pop_per_km2, percentile_density, perc_essential, percentile_essential, sum_score_decile, sum_percentile_score, SIMD2020v2_Decile) %>%
  arrange(desc(sum_percentile_score)) %>%
  clean_names() %>%
  gt() %>%
  fmt_number(columns = vars(pop_per_km2, perc_essential), decimals = 0) %>%
  tab_options(container.height = 400)
```


Sources
=====================================

Row
-------------------------------------
###

This work was inspired by the Colorado Social Distancing Index created by the Colorado Health Institute
https://www.coloradohealthinstitute.org/research/colorado-covid-19-social-distancing-index

Overcrowding data from Scottish Index of Multiple Deprivation 2020 (SIMD)
https://www.gov.scot/collections/scottish-index-of-multiple-deprivation-2020/

Population density from the 2011 Scotland Data Zones
Data and shapefiles from https://data.gov.uk/dataset/ab9f1f20-3b7f-4efa-9bd2-239acf63b540/data-zone-boundaries-2011

Occupations from the 2011 Census
https://www.scotlandscensus.gov.uk/ods-web/data-warehouse.html
File QS605SC

All data and analytic code are at https://github.com/JessButler/social_distancing


